some title

### 2018-02-04 2018/chapters/2_regression.tex
chi-sq def

Now we show that the~latter regression produces $\hat \beta_1$ coefficient which
is exatly the coefficient from the~`one-step' regression of~original $y$ onto
original $x$ and $z$. Recall that the~vector $y$ can be split~up into a~sum of
some multiple of $x$ and some multiple of $z$. Since the second term is
the~orthogonal component its projection yields zero. The~multiple of $x$
is equal to $\hat \beta_1$ by construction.

Assume that the coefficient at $\tilde x$ is some unknown variable $\hat \delta$.
Then consider the similar triangles in the $Lin(x,z)$. From the~proportions
we obtain:
\[
\frac{CE}{CA} = \frac{CD}{CB} \Leftrightarrow \frac{\hat \beta_1 x}{x} = \frac{\hat \delta \tilde x}{\tilde x} \Rightarrow \hat \beta_1 = \hat \delta
\]

>>>>

Now we show that the~latter regression produces $\hat \beta_1$ coefficient which
is exatly the~coefficient from the~`one-step' regression of~original $y$ onto
original $x$ and $z$. Recall that the~vector $y$ can be split~up into a~sum of
some multiple of $x$ and some multiple of $z$. Since the second term is
the~orthogonal component its projection yields zero. The~multiple of $x$
is equal to $\hat \beta_1$ by construction.

Assume that the~coefficient at~$\tilde x$ is some unknown variable $\hat \delta$.
Then consider the~similar triangles in the $Lin(x,z)$. From the~proportions
we obtain:
\[
\frac{CE}{CA} = \frac{CD}{CB} \Leftrightarrow \frac{\hat \beta_1 x}{x} = \frac{\hat \delta \tilde x}{\tilde x} \Rightarrow \hat \beta_1 = \hat \delta
\]

### 2018-02-04 2018/chapters/4_distributions.tex
restructuring finished, fwl v2
chi-sq def

Consider a random vector $z \in \mathbb{R}^n$ which components are independent
and follow standard normal distribution, $z_i \sim \mathcal{N}(0,1)$.
Consider also a fixed $k$-dimensional subspace in $\mathbb{R}^n$ $L$.
Let the projection of vector $z$ onto the subspace $L$ be $\hat z$ and
its length squared $Q$
\[
Q = \lVert \hat z \rVert^2 = \langle \hat z, \hat z \rangle = \hat z^T \hat z
\]
Then $Q$ follows the chi-squared distribution with $k$ degrees of freedom.

@@

\begin{figure}[ht!]
\begin{center}
\subfigure[]{
\includegraphics[width=0.45\linewidth]{figures/ttest.pdf}
\label{fig:ttest_3d}}
%\hspace{4ex}
\subfigure[]{
\includegraphics[width=0.45\linewidth]{figures/ttest_lin.pdf}
\label{fig:ttest_lin}}
\caption{\subref{fig:pcorr_t_x}: Regression of $y$ onto $Lin(x, \mathbf{1})$ and appropriate projections;
\subref{fig:pcorr_t_y}: $Lin^{\perp}(\mathbf{1})$.}
\end{center}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.55\linewidth]{figures/ftest.pdf}
\caption{F-statistic as the cotangens squared of $\varphi$}
\label{fig:ftest}
\end{figure}

>>>>

\marginnote{
Let $z_i \stackrel{iid}{\sim} \mathcal{N}(0,1)$.
Then $Q$ follows the chi-squared distribution with $k$ degrees of freedom
if it can be written as
\[
Q = z_1^2 + z_2^2 + \ldots + z_k^2.
\]

This definition is a particular case of the geometric one.
Consider projecting a vector $z = (z_1, z_2, \ldots, z_n)$ form $\mathbb{R}^n$
onto the~$k$-dimensional subspace $S$ of vectors which first $k$ coordinates
are arbitrary and all the rest are zeros. As a result we would get
\[
\hat z = (z_1, z_2, \ldots, z_k, 0, \ldots, 0).
\]
Squaring the length of the projection, we obtain
\[
Q = \lVert \hat z \rVert = z_1^2 + z_2^2 + \ldots + z_k^2.
\]
}

\begin{theorem}
Consider a random vector $z \in \mathbb{R}^n$ which components are independent
and follow standard normal distribution, $z_i \sim \mathcal{N}(0,1)$.
Consider also a fixed $k$-dimensional subspace $L$ in $\mathbb{R}^n$.
Let the projection of vector $z$ onto the subspace $L$ be $\hat z$ and
its length squared $Q$
\[
Q = \lVert \hat z \rVert^2 = \langle \hat z, \hat z \rangle = \hat z^T \hat z
\]
Then $Q$ follows the chi-squared distribution with $k$ degrees of freedom.
\end{theorem}

\begin{proof}
First, it can be shown that the~projected vector $\hat z$ is the~original vector $z$
multiplied by the~projection matrix $H = X(X^T X)^{-1}X^T$ where the columns of $X$
are fixed linearly independent vectors $x_1, \ldots, x_k$ in $L$
or equivalently $colX = Lin(x_1, \ldots, x_k)$.
This matrix is  also often referred to as `hat-matrix'.
Then the statement in~the~theorem can be rewritten as follows:
\[
\hat z^T \hat z = (Hz)^T Hz = z^T H^T H z = z^T H^2 z = z^T H z,
\]
applying the idempotence property in the last step.

Another nice property of~the~hat-mattix is symmetry.
Thus, it can be decomposed as
\[
H = P D P^T,
\]
where we choose the vectors of matrix $P$ to be unit and orthogonal,
and $D = diag{\lambda}$ where $\lambda_i$ is an eigenvalue of $H$.

\begin{marginfigure}
  \includegraphics[width=\linewidth]{figures/04_chi_squared_example.pdf}
  \caption{Consider a $3$-dimensional example, $colX = Lin(x_1, x_2)$ and $col^{\perp}X = Lin(x_3)$.
  $H x_1 = x_1$ and $H x_2 = x_2$ since they are in $colX$. However, $H x_3 = 0$ as $x_3 \perp colX$.
  Projecting an arbitrary vector onto $colX$ yileds $Hy = \hat y \in Lin(x_1, x_2)$
  while projecting onto $col^{\perp}X$ results in $(I-H)y = \hat{\hat y} \in Lin(x_3)$.}
\end{marginfigure}

Since $H^2 = H$ the eigenvalues are either $0$ or $1$.
Recall that $H$ projects a vector onto $colX$.
Then for any $x_i$, $i = 1, \ldots, k$, $H x_i = x_i \cdot 1$ since
any $x_i$ is already in $colX$. This implies that $\lambda_1 = \ldots = \lambda_k = 1$.
There are also $n-k$ vectors in the subspace orthogonal to $colX$.
So for any $x_i$, $i= k+1, \ldots, n$, the orthogonal projection yields zero.
We conclude that $\lambda_{k+1} = \ldots = \lambda_n = 0$.

Rewritting the theroem statement further, we obtain
\[
z^T H z = z^T P D P^T z = (P^T z)^T D (P^T z) = \tilde z^T D \tilde z = \tilde z_1^2 + \ldots + \tilde z_k^2.
\]
Now we explore $\tilde z$ given $z \sim \mathcal{N}(0, I)$:
\begin{align*}
&\tilde z = P^T z \\
&\E(\tilde z) = \E(P^T z) = P^T \E(z) = 0 \\
&\Var(\tilde z) = \Var(P^T z) = P^T \Var(z) (P^T)^T = P^T P = I
\end{align*}
So we conclude that $\tilde z_1^2 + \ldots + \tilde z_k^2 \sim \chi^2_k$.

\end{proof}

@@

\begin{figure}[ht!]
\begin{center}
\subfigure[]{
\includegraphics[width=0.45\linewidth]{figures/04_ttest.pdf}
\label{fig:ttest_3d}}
%\hspace{4ex}
\subfigure[]{
\includegraphics[width=0.45\linewidth]{figures/04_ttest_lin.pdf}
\label{fig:ttest_lin}}
\caption{\subref{fig:pcorr_t_x}: Regression of $y$ onto $Lin(x, \mathbf{1})$ and appropriate projections;
\subref{fig:pcorr_t_y}: $Lin^{\perp}(\mathbf{1})$.}
\end{center}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.55\linewidth]{figures/04_ftest.pdf}
\caption{F-statistic as the cotangens squared of $\varphi$}
\label{fig:ftest}
\end{figure}
